The great codivergence bake off

### 2025-11-26 manuscript/ms.tex
Working on manuscript.
Sketching out an introduction and adding some text about
the nature of the simulations.

@@

@@

>>>>

To understand the distribution of biodiversity, the field of historical
biogeography considers how past environmental changes have shaped diversity
within and among species.
Environmental processes that affect whole biological communities can cause
speciation or extinction across multiple species at the same time.
Biogeographers are often interested in testing for patterns of temporally
clustered speciation events as evidence that a shared process caused multiple
lineages to diverge.
For example, finding that a number of evolutionary lineages diverged
simultaneously can be difficult to explain via species-specific processes alone and
provides compelling evidence for a shared process of diversification, like the
fragmentation of an environment.

One way to test the predictions of such processes of diversification is to
infer the temporal pattern of divergences across multiple taxa, and determine
whether any subsets of the taxa shared the same divergence times.
Methods have been developed to test for such patterns of shared divergence
times across taxa using approximate Bayesian computation (ABC)
\citep{Hickerson2006,Huang2011,Oaks2014dpp}
and likelihood-based Bayesian inference
\citep{Oaks2018ecoevolity,Oaks2019codemog}.
Both approaches use Bayesian model choice to infer the temporal pattern of
divergences across multiple pairs of populations or species.

Several of the ABC methods were primarily developed for analyzing a small
number of unlinked genetic loci each comprising linked, nonrecombining sites.
These methods often struggle to detect multiple divergence times across pairs of
populations \citep{Oaks2012, Oaks2014reply} or have little information to
update \textit{a priori} expectations \citep{Oaks2014dpp}.
More fundamentally, the loss of information inherent to ABC approaches can
prevent them from discriminating among models
\citep{Robert2011,Marin2014,Green2015}.

By more efficiently using information from a large number of effectively
unlinked sites from across the genomes of sampled organisms, the
full-likelihood Bayesian methods enable more accurate and precise estimates of
shared divergneces
\citep{Oaks2018ecoevolity,Oaks2019codemog}.

This raises a question.
Given a dataset of one or a small number of loci of linked sites,
is it better to use an ABC approach that correctly models such loci, but
but suffers from low power and bias?
Or, is it better to use a full-likelihood approach that has more power,
but incorrectly assumes the sites in the data are unlinked?
\jrocomment{Need to flesh out this intro more.}

@@

Ideally, we would have applied both the full-likelihood and ABC methods
to the datasets simulated using \pymsbayes.
However, this was not possible, because \dppmsbayes does not store the full
simulated sequence data, only the population genetic summary statistics used by
the ABC algorithm.

@@

As mentioned in the methods,
it would have been ideal to apply both the full-likelihood and ABC methods
to the data simulated via \dppmsbayes.
Because \dppmsbayes only stores summary statistics from the genetic data, this
was not possible.
However, it is unlikely that differences between the simulators implemented in
\dppmsbayes and \ecoevolity could explain the differences in performance
between full-likelihood and ABC inference methods.
Both simulators have been independently validated by suites of tests in the
\pymsbayes and \ecoevolity distributions.
Also, the assumption that the simulators are producing nearly identically
distributed data (barring minor differences in parameterizations between the
implementations) is supported by our finding that both inference methods
perform very similarly when the size of the simulated data sets are small
(i.e., one 500-site locus) and diverge in performance as we increased the size
of the simulated data sets (\figs \ref{fig:modelerror} \&
\ref{fig:divtimeerror}).
